---
title: "Reproducible Research: Peer Assessment 1"

author : "**Ladan Rabbani**"

date:     "**Aug 2015**"
  
---
## Introduction
 This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.
 Through this report you can see that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport.

## Prepare the R markdown file 

First, we set echo equal a TRUE and results equal a 'hold' as global options for this document.
```{r}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

## Loading and preprocessing the data
load the required packages :
```{r}
library(data.table)
library(ggplot2)
library(dplyr)
```
read the data :

```{r}
activity <- read.csv("activity.csv", sep = ",",colClasses = c("numeric","character","numeric"))
```

preprocess the data :

```{r}
activity.date <- as.Date(activity$date, format = "%Y-%m-%d")
activity.interval <- as.factor(activity$interval)
```
now ,lets check the data using str()method just to look at the data and variables :
```{r}
str(activity)
```

## What is mean total number of steps taken per day?
for this part of assignment ,we ignore the missing value in the data set .

1.first calculate the total number of steps taken per day :
```{r}
steps.day <- aggregate(steps~date, data =activity, FUN = sum, na.rm= "TRUE")

colnames(steps.day) <-c("date","steps")

head(steps.day)
```

2. now make the histogram of the total number of steps taken per day ;
```{r}
ggplot(steps.day, aes(x=steps))+
  geom_histogram(fill="red", binwidth=800)+
  labs(title= "Total Number Of Steps Each Day ",
       x= "Number of steps")+
  theme_bw()
```

3. Calculate and report the mean and median of the total number of steps taken per day :

```{r}
steps.day.mean <- mean(steps.day$steps)
steps.day.median <- median(steps.day$steps)

steps.day.mean
steps.day.median
```

## What is the average daily activity pattern?
first calculate the aggregation of steps by 5 minutes interval :

```{r}
interval.step <- aggregate(steps~interval, data = activity, FUN = mean, na.rm="TRUE")
```

1.make a time series plot of 5minutes interval and average number of step taken, averaged across all days :

```{r}
ggplot(interval.step, aes(x=interval,y=steps))+
  geom_line(color="green",size=0.8)+
  labs(title="Average Daily Activity",x="interval",y="number of steps")+
  theme_bw()
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max.interval.step <- interval.step[which.max(interval.step$step),]$interval
```

## Imputing missing values

1. first we Calculate and report the total number of missing values in the dataset :
```{r}
Miss.NA <-sum(is.na(activity$steps))

Miss.NA
```

The total number of missing value is 2304 .

2.Devise a strategy for filling in all of the missing values in the dataset. 

  we can use the means for 5 minutes interval as fillers for missing items ;

3. create a new dataset that is equal to the original dataset but with the missing data filled in :

```{r}
activity<- merge(activity, interval.step, by="interval", suffixes = c("", ".y"))
 
S.NA <- is.na(activity$steps)
activity$steps[S.NA] <- activity$steps.y[S.NA]

activity <- activity[,c(1:3)]
```
now just to make sure there are no missing values remaining :
```{r}
sum(is.na(activity$steps))
```
  *zero* shows that there are *no missing value * in the data set .
  
3.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r}
steps.day.noNA <- aggregate(steps ~ date, data = activity, FUN = sum)

ggplot(steps.day.noNA, aes(x=steps))+
  geom_histogram(fill="dark orange", binwidth=800)+
  labs(title="Number Of Steps Each Day (missing value filled)",
       x="Number of steps each day" )+
  theme_bw()
```

*calculate and report the **mean** and **median** total number of steps taken per day* :

```{r}
steps.day.noNA.mean<- mean(steps.day.noNA$steps)
steps.day.noNa.median <- median(steps.day.noNA$steps)
```

**Do these values differ from the estimates from the first part of the assignment? **
the mean value stayed the same , the median is very slightly different 

* **Before filling NA's** 
      + mean : 10766.189
      + median: 10765
      
* **After filling NA's**
      + mean : 10766.189
      + median: 10766.189
      
**What is the impact of imputing missing data on the estimates of the total daily number of steps?**

its look like the median value has shifted towards mean value but by looking at both histograms its look like having the missing value in the data will not have a big impact on our prodection and it doesnt have a negative affect . 
      
## Are there differences in activity patterns between weekdays and weekends?

we will use the dateset with the filled in missing value for this part .

1.Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
day.type <- function(date)  {
  if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday"))
    {
    "weekend"
     } else 
       {
    "weekday"
       }
    }
activity$day.type<- as.factor(sapply(activity$date, day.type))
```

2.Make a panel plot containing a time series plot of the 5-minute interval  and the average number of steps taken, averaged across all weekday days or weekend days. 

```{r}

par(mfrow = c(2, 1))
 for (type in c("weekend", "weekday")) {
   steps.day.type <- aggregate(steps~interval, data = activity, subset=activity$day.type == type, FUN =mean)
   plot(steps.day.type, type = "l", main = type, col="blue", lwd=2)
 }
```


